openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2023-05-23 2023-05-revision-1/main.tex
paper: update text and table 1 with FIt-SNE pypi/conda checkmarks

Easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques. Despite the many recent theoretical advancements, popular t-SNE libraries have slowly incorporated them into their implementations.
In \proglang{Python}, the most commonly used implementation of the t-SNE algorithm comes from \pkg{scikit-learn}~\citep{pedregosa2011scikit}, which is nicely integrated with \proglang{Python} data science environment but suffers from poor scalability. Other implementations from the \proglang{R}~\citep{team2013r,krijthe2015rtsne} and \proglang{Julia}~\citep{bezanson2017julia,julia_tsne} programming languages also scale poorly (see Sec.~\ref{s:speed}). \pkg{MulticoreTSNE}~\citep{Ulyanov2016} enjoys better scalability but is likewise limited when applied to larger data sets. Additionally, none of these implementations include recently proposed extensions that can result in more globally consistent embeddings. \pkg{FIt-SNE}~\citep{linderman2019fast} alleviates these problems and provides bindings to both the \proglang{Python} and \proglang{R} programming languages. However, \pkg{FIt-SNE} is difficult to install and distribute and provides an API largely inconsistent with the \proglang{Python} data science ecosystem. None of the existing t-SNE implementations support adding new samples into existing embeddings. To alleviate all these problems, and to support newly proposed additions to t-SNE algorithms, we here present \opentsne, an open-source \proglang{Python} implementation of the t-SNE algorithm. \opentsne is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers. It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.

@@

\toprule
\textbf{Ease of installation} \\
Native \proglang{Python} installation & \checkmark & & & \checkmark & & \\
\pkg{PyPI} package & \checkmark & \checkmark & & \checkmark & & \\
\pkg{conda} package & \checkmark & & & \checkmark & & \\

@@

\hline
\textbf{Advanced features and extensions} \\
%Multiscale Gaussian kernels & & & & \checkmark & \checkmark & & \\
Extensible affinity models & & & & \checkmark & & \\
Variable degrees of freedom & & & \checkmark & \checkmark & & \\
Variable exaggeration & & & \checkmark & \checkmark & & \\
Globally-consistent initialization & & & \checkmark & \checkmark & & \\
Automatic learning rate & & & \checkmark & \checkmark & & \\
Adding samples to embeddings & & & & \checkmark & & \\
Callback system & & & & \checkmark & & \\
Interactive optimization  & & & & \checkmark & & \\

>>>>

Easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques. Despite the many recent theoretical advancements, popular t-SNE libraries have slowly incorporated them into their implementations.
In \proglang{Python}, the most commonly used implementation of the t-SNE algorithm comes from \pkg{scikit-learn}~\citep{pedregosa2011scikit}, which is nicely integrated with \proglang{Python} data science environment but suffers from poor scalability. Other implementations from the \proglang{R}~\citep{team2013r,krijthe2015rtsne} and \proglang{Julia}~\citep{bezanson2017julia,julia_tsne} programming languages also scale poorly (see Sec.~\ref{s:speed}). \pkg{MulticoreTSNE}~\citep{Ulyanov2016} enjoys better scalability but is likewise limited when applied to larger data sets. Additionally, none of these implementations include recently proposed extensions that can result in more globally consistent embeddings. \pkg{FIt-SNE}~\citep{linderman2019fast} alleviates these problems and provides bindings to both the \proglang{Python} and \proglang{R} programming languages. However, none of the existing t-SNE implementations support adding new samples into existing embeddings. To alleviate all these problems, and to support newly proposed additions to t-SNE algorithms, we here present \opentsne, an open-source \proglang{Python} implementation of the t-SNE algorithm. \opentsne is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers. It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.

@@

\toprule
\textbf{Ease of installation} \\
\pkg{PyPI} package & \checkmark & \checkmark & \checkmark & \checkmark & & \\
\pkg{conda} package & \checkmark & & \checkmark & \checkmark & & \\

@@

\hline
\textbf{Advanced features and extensions} \\
Extensible affinity models & & & & \checkmark & & \\
Variable degrees of freedom & & & \checkmark & \checkmark & & \\
Variable exaggeration & & & \checkmark & \checkmark & & \\
Globally-consistent initialization & & & \checkmark & \checkmark & & \\
Automatic learning rate & & & \checkmark & \checkmark & & \\
Adding samples to embeddings & & & & \checkmark & & \\
Callback system & & & & \checkmark & & \\
Interactive optimization  & & & & \checkmark & & \\
